A performance comparison of Dask and Apache Spark for data-intensive neuroimaging pipelines

### 2019-07-29 latex/paper.tex
Renamed framework to engine

In the past few years, neuroimaging has entered the Big Data era due to the joint
increase in image resolution, data sharing, and study sizes. However, no particular
Big Data engines has emerged in this field, and several alternatives remain
available. We compare two popular Big Data engines with Python APIs, Apache Spark and
Dask, for their runtime performance in processing neuroimaging pipelines. Our
evaluation uses two synthetic pipelines processing the \SI{81}{\giga\byte} Big Brain
image, and a real pipeline processing anatomical data from more than 1,000 subjects.
We benchmark these pipelines using various combinations of task durations, data
sizes, and numbers of workers, deployed on a 8-node (8 cores ea.) compute cluster in
Compute Canada's Arbutus cloud. We evaluate PySpark's RDD API against Dask's Bag,
Delayed and Futures. Results show that despite slight differences between Spark and
Dask, both frameworks perform comparably. However, Dask pipelines risk being limited
by Python's GIL depending on task type and cluster configuration. In all cases, the
major limiting factor was data transfer. While either framework is suitable for
neuroimaging pipelines, more effort needs to be placed in reducing data transfer time.
\end{abstract}

@@

\section{Introduction}
% Context
The recent rise in data sharing and improved data collection strategies
have brought neuroimaging to the Big Data era ~\cite{ALFAROALMAGRO:18,
van2014human}. Existing neuroimaging workflow engines, such as
Nipype~\cite{Nipype:11}, are well suited for processing the standard
compute-intensive neuroimaging pipelines, but lack incorporated Big Data
strategies (i.e. in-memory computing, data locality and lazy-evaluation) to
improve performance of the increasingly prevalent data-intensive pipelines.
As was noted in ~\cite{hayot2019performance}, in-memory computing, coupled
with data locality, can bring significant performance improvements to data
intensive neuroimaging pipelines. We extend this work by studying the
differences between two Big Data frameworks, Dask~\cite{Dask:15} and Apache
Spark~\cite{Spark:16}, for their suitability in the processing of neuroimaging
pipelines. Our goal is to test whether Spark or Dask has a clear
performance advantage to process Big neuroimaging Data.

%Similarities
Spark and Dask both offer in-memory computing, data locality, and lazy evaluation,
which is common for Big Data frameworks. Both their schedulers operate dynamically,
which is good when task runtimes are not known ahead of time~\cite{Dask:15}. They
also provide rich, high-level programming APIs, and support a variety of
infrastructure schedulers, such as Mesos~\cite{hindman2011mesos}, YARN~\cite{vavilapalli2013apache}, or HPC clusters (Dask only). Over
these similarities, the frameworks have differences.

First and foremost, Spark is written in Scala while Dask is in Python.
Given the popularity of Python in scientific communities, this arguably
gives an edge to Dask due to data serialization costs from Python to Scala.
On the other hand, Python's Global Interpreter Lock (GIL) might reduce
parallelism in some cases. This difference in programming languages also
has qualitative implications. As part of the \href{http://scipy.org}{SciPy}
ecosystem, Dask provides almost transparent parallelization of applications
manipulating Numpy arrays or Pandas data frames. On the other hand, Spark's
Java, R and Python APIs allow to easily parallelize analyses that combine
these languages, with reduced performance loss. Our study focuses on
performance, although we recognize that other factors will also impact the
choice of a framework in practice.

@@

\subsubsection{Apache Spark} Apache Spark is a widely-used 
general-purpose Big Data engine.
Its main abstraction, the Resilient Distributed Dataset
(RDD)~\cite{RDD}, is a fault-tolerant, parallel collection of data elements. It 
achieves fault tolerance through the recording of data lineage, the
sequence of operations used to modify the original data. The RDD is the basis
of Spark's other data structures, namely, Datasets
and DataFrames. Datasets are similar to RDDs but additionally use Spark SQL's
optimized execution to further improve performance. DataFrames, used to process
tabular data, are Datasets where the data is organized into named-columns.
While the DataFrame API exists in all the available language APIs, Datasets are
limited to Scala and Java.

@@

The Dask framework is compatible with numerous distributed schedulers, including 
\href{https://github.com/dask/dask-yarn}{YARN} and
\href{https://github.com/mrocklin/dask-mesos}{Mesos}, similarly to Spark. Dask 
also provides its own distributed scheduler, known as the Dask Distributed
scheduler. Although Dask is compatible with schedulers commonly found in HPC
clusters, we chose to use Dask's Distributed scheduler, to keep the environment
balanced between both frameworks.

>>>>

In the past few years, neuroimaging has entered the Big Data era due to the joint
increase in image resolution, data sharing, and study sizes. However, no particular
Big Data engines has emerged in this field, and several alternatives remain
available. We compare two popular Big Data engines with Python APIs, Apache Spark and
Dask, for their runtime performance in processing neuroimaging pipelines. Our
evaluation uses two synthetic pipelines processing the \SI{81}{\giga\byte} Big Brain
image, and a real pipeline processing anatomical data from more than 1,000 subjects.
We benchmark these pipelines using various combinations of task durations, data
sizes, and numbers of workers, deployed on a 8-node (8 cores ea.) compute cluster in
Compute Canada's Arbutus cloud. We evaluate PySpark's RDD API against Dask's Bag,
Delayed and Futures. Results show that despite slight differences between Spark and
Dask, both engines perform comparably. However, Dask pipelines risk being limited
by Python's GIL depending on task type and cluster configuration. In all cases, the
major limiting factor was data transfer. While either engine is suitable for
neuroimaging pipelines, more effort needs to be placed in reducing data transfer time.
\end{abstract}

@@

\section{Introduction}
% Context
The recent rise in data sharing and improved data collection strategies
have brought neuroimaging to the Big Data era ~\cite{ALFAROALMAGRO:18,
van2014human}. Existing neuroimaging workflow engines, such as
Nipype~\cite{Nipype:11}, are well suited for processing the standard
compute-intensive neuroimaging pipelines, but lack incorporated Big Data
strategies (i.e. in-memory computing, data locality and lazy-evaluation) to
improve performance of the increasingly prevalent data-intensive pipelines.
As was noted in ~\cite{hayot2019performance}, in-memory computing, coupled
with data locality, can bring significant performance improvements to data
intensive neuroimaging pipelines. We extend this work by studying the
differences between two Big Data engines, Dask~\cite{Dask:15} and Apache
Spark~\cite{Spark:16}, for their suitability in the processing of neuroimaging
pipelines. Our goal is to test whether Spark or Dask has a clear
performance advantage to process Big neuroimaging Data.

%Similarities
Spark and Dask both offer in-memory computing, data locality, and lazy evaluation,
which is common for Big Data engines. Both their schedulers operate dynamically,
which is good when task runtimes are not known ahead of time~\cite{Dask:15}. They
also provide rich, high-level programming APIs, and support a variety of
infrastructure schedulers, such as Mesos~\cite{hindman2011mesos}, YARN~\cite{vavilapalli2013apache}, or HPC clusters (Dask only). Over
these similarities, the engines have differences.

First and foremost, Spark is written in Scala while Dask is in Python.
Given the popularity of Python in scientific communities, this arguably
gives an edge to Dask due to data serialization costs from Python to Scala.
On the other hand, Python's Global Interpreter Lock (GIL) might reduce
parallelism in some cases. This difference in programming languages also
has qualitative implications. As part of the \href{http://scipy.org}{SciPy}
ecosystem, Dask provides almost transparent parallelization of applications
manipulating Numpy arrays or Pandas data frames. On the other hand, Spark's
Java, R and Python APIs allow to easily parallelize analyses that combine
these languages, with reduced performance loss. Our study focuses on
performance, although we recognize that other factors will also impact the
choice of an engine in practice.

@@

\subsubsection{Apache Spark} Apache Spark is a widely-used 
general-purpose Big Data engine.
Its main abstraction, the Resilient Distributed Dataset
(RDD)~\cite{RDD}, is a fault-tolerant, parallel collection of data elements. It 
achieves fault tolerance through the recording of data lineage, the
sequence of operations used to modify the original data. The RDD is the basis
of Spark's other data structures, namely, Datasets
and DataFrames. Datasets are similar to RDDs but additionally use Spark SQL's
optimized execution engine to further improve performance. DataFrames, used to process
tabular data, are Datasets where the data is organized into named-columns.
While the DataFrame API exists in all the available language APIs, Datasets are
limited to Scala and Java.

@@

The Dask engine is compatible with numerous distributed schedulers, including 
\href{https://github.com/dask/dask-yarn}{YARN} and
\href{https://github.com/mrocklin/dask-mesos}{Mesos}, similarly to Spark. Dask 
also provides its own distributed scheduler, known as the Dask Distributed
scheduler. Although Dask is compatible with schedulers commonly found in HPC
clusters, we chose to use Dask's Distributed scheduler, to keep the environment
balanced between both engines.
